---
title: "Exploring Storms"
author: "Hyder Ali"
execute:
  echo: false
format:
  html:
    theme:
      light: cosmo
      dark: darkly
    toc: true
    toc-title: "Contents"
    embed-resources: true
---

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages(library(tidyverse))
```

# Introduction

In this report, we analyze the historical hurricane data from the NOAA Atlantic hurricane database (`dplyr::storms`). This dataset spans from 1975 to 2024 and tracks storm locations and characteristics.

To understand trends in storm activity, we focus on **named storms** (tropical storms and hurricanes given official names), excluding unnamed tropical depressions (which are recorded using generic codes like "ALxxYYYY").

# Named Storms per Year

The visualization below illustrates the annual count of unique named storms in the Atlantic basin. 

```{r}
#| label: plot-named-storms
#| fig-cap: "Annual counts of named storms in the Atlantic Basin (1975-2024)."
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 5.5

# Filter out unnamed/numbered depressions and calculate distinct storms per year
named_storms <- storms %>%
  filter(!grepl("^(AL|Al)[0-9]", name)) %>%
  distinct(year, name) %>%
  count(year, name = "num_storms")

# Calculate summary stats for annotations
avg_storms <- mean(named_storms$num_storms)
peak_year <- named_storms %>% 
  slice_max(num_storms, n = 1, with_ties = FALSE)

# Generate a highly polished, interactive-feeling visualization
ggplot(named_storms, aes(x = year, y = num_storms)) +
  # Reference line for the historical average
  geom_hline(yintercept = avg_storms, linetype = "dashed", color = "#94a3b8", size = 0.8) +
  
  # Smooth gradient area underneath the line for modern aesthetics
  geom_area(fill = "#818cf8", alpha = 0.15) +
  
  # The trend line
  geom_line(color = "#4f46e5", size = 1.2, alpha = 0.85) +
  
  # Data points
  geom_point(color = "#4f46e5", size = 2.5, alpha = 0.9) +
  
  # Highlight the peak year with a custom circle
  geom_point(data = peak_year, aes(x = year, y = num_storms), 
             color = "#ef4444", size = 4.5, stroke = 2, shape = 21, fill = "white") +
  
  # Add dynamic labels for the peak year and average line
  geom_text(data = peak_year, 
            aes(x = year - 1, y = num_storms + 1, 
                label = paste0("Peak Year: ", year, " (", num_storms, " storms)")),
            color = "#ef4444", fontface = "bold", size = 3.8, hjust = 1) +
  
  annotate("text", x = 1976, y = avg_storms + 0.9, 
           label = paste0("Historical Average: ", round(avg_storms, 1)),
           color = "#475569", fontface = "italic", size = 3.5, hjust = 0) +
  
  # Labels and scales
  labs(
    title = "Annual Number of Named Storms (1975 - 2024)",
    subtitle = "Analysis of NOAA Atlantic hurricane database",
    x = "Year",
    y = "Number of Named Storms",
    caption = "Source: NOAA HURDAT database (dplyr::storms)"
  ) +
  scale_x_continuous(breaks = seq(1975, 2025, by = 5)) +
  scale_y_continuous(breaks = seq(0, 35, by = 5), limits = c(0, 35)) +
  
  # Theme customization
  theme_minimal(base_family = "sans") +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#1e293b", margin = margin(b = 4)),
    plot.subtitle = element_text(size = 11, color = "#64748b", margin = margin(b = 15)),
    plot.caption = element_text(size = 8, color = "#94a3b8", hjust = 1, margin = margin(t = 15)),
    axis.title.x = element_text(face = "bold", size = 11, color = "#475569", margin = margin(t = 10)),
    axis.title.y = element_text(face = "bold", size = 11, color = "#475569", margin = margin(r = 10)),
    axis.text = element_text(color = "#64748b", size = 9.5),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(color = "#e2e8f0", size = 0.5),
    panel.grid.minor.y = element_blank(),
    plot.margin = margin(20, 20, 20, 20)
  )
```

# Storm Activity Table

For details, here are the counts of named storms for the most recent 10 years of data:

```{r}
#| label: storm-table
#| message: false

named_storms %>%
  filter(year >= 2015) %>%
  arrange(desc(year)) %>%
  knitr::kable(
    col.names = c("Year", "Number of Named Storms"),
    align = "cr",
    caption = "Named Storm Counts (2015-2024)"
  )
```
