---
title: "Exploring Storms"
author: "Hyder Ali"
execute:
  echo: false
format:
  html:
    theme:
      light: cosmo
      dark: darkly
    toc: true
    toc-title: "Contents"
    embed-resources: true
---

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages(library(tidyverse))
```

# Introduction

In this report, we analyze the historical hurricane data from the NOAA Atlantic hurricane database (`dplyr::storms`). This dataset spans from 1975 to 2024 and tracks storm locations and characteristics.

To understand trends in storm activity, we focus on three major storm types: **tropical depressions**, **tropical storms**, and **hurricanes** (which can be named or unnamed).

# Storm Activity by Type

The visualization below illustrates the annual count of unique storms in the Atlantic basin, categorized by their storm type.

```{r}
#| label: plot-named-storms
#| fig-cap: "Annual counts of storms by type in the Atlantic Basin (1975-2024)."
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 5.5

# Filter out unnamed/numbered depressions and calculate distinct storms per year for the table
named_storms <- storms %>%
  filter(!grepl("^(AL|Al)[0-9]", name)) %>%
  distinct(year, name) %>%
  count(year, name = "num_storms")

# Filter to tropical depressions, tropical storms, and hurricanes, and count distinct storms per year/status
storm_types <- storms %>%
  filter(status %in% c("tropical depression", "tropical storm", "hurricane")) %>%
  mutate(status = factor(status, 
                         levels = c("tropical depression", "tropical storm", "hurricane"),
                         labels = c("Tropical Depression", "Tropical Storm", "Hurricane"))) %>%
  distinct(year, name, status) %>%
  count(year, status, name = "count") %>%
  complete(year, status, fill = list(count = 0))

# Generate a highly polished, interactive-feeling stacked area chart
ggplot(storm_types, aes(x = year, y = count, fill = status)) +
  # Stacked area chart with thin white borders for visual clarity
  geom_area(color = "white", size = 0.25, alpha = 0.85) +
  
  # Custom color scale: light blue -> orange -> dark red
  scale_fill_manual(
    values = c(
      "Tropical Depression" = "#93c5fd",
      "Tropical Storm" = "#fb923c",
      "Hurricane" = "#b91c1c"
    )
  ) +
  
  # Labels and scales
  labs(
    title = "Annual Storm Activity by Type (1975 - 2024)",
    subtitle = "Evolution of tropical depressions, tropical storms, and hurricanes in the Atlantic basin",
    x = "Year",
    y = "Number of Storms",
    fill = "Storm Type",
    caption = "Source: NOAA HURDAT database (dplyr::storms)"
  ) +
  scale_x_continuous(breaks = seq(1975, 2025, by = 5), expand = c(0, 0)) +
  scale_y_continuous(breaks = seq(0, 80, by = 10), limits = c(0, 80), expand = c(0, 0)) +
  
  # Theme customization
  theme_minimal(base_family = "sans") +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#1e293b", margin = margin(b = 4)),
    plot.subtitle = element_text(size = 11, color = "#64748b", margin = margin(b = 15)),
    plot.caption = element_text(size = 8, color = "#94a3b8", hjust = 1, margin = margin(t = 15)),
    axis.title.x = element_text(face = "bold", size = 11, color = "#475569", margin = margin(t = 10)),
    axis.title.y = element_text(face = "bold", size = 11, color = "#475569", margin = margin(r = 10)),
    axis.text = element_text(color = "#64748b", size = 9.5),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_line(color = "#e2e8f0", size = 0.5),
    panel.grid.minor.y = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(face = "bold", size = 10, color = "#475569"),
    legend.text = element_text(size = 9, color = "#64748b"),
    plot.margin = margin(20, 20, 20, 20)
  )
```

# Storm Activity Table

For details, here are the counts of named storms for the most recent 10 years of data:

```{r}
#| label: storm-table
#| message: false

named_storms %>%
  filter(year >= 2015) %>%
  arrange(desc(year)) %>%
  knitr::kable(
    col.names = c("Year", "Number of Named Storms"),
    align = "cr",
    caption = "Named Storm Counts (2015-2024)"
  )
```
